---
title: "Analisis de Componentes principales"
author: Gonzalo J. Ale Pezo
institute: "DEPARTAMENTO DE MATEMÁTICA Y ESTADÍSTICA, UNSAAC"
date:  "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, cho = FALSE, warning = FALSE, message = FALSE)
```

<style>
body {
text-align: justify}
</style>

# PCA Overview

- PCA is used in exploratory data analysis and for making decisions in predictive models.

- PCA commonly used for dimensionality reduction by using each data point onto only the first few principal components (most cases first and second dimensions) to obtain lower-dimensional data while keeping as much of the data’s variation as possible.

- The first principal component can equivalently be defined as a direction that maximizes the variance of the projected data.

- The principal components are often analyzed by eigendecomposition of the data covariance matrix or singular value decomposition (SVD) of the data matrix.

- Reducing the number of variables from a data set naturally leads to inaccuracy, but the trick in the dimensionality reduction is to allow us to make correct decisions based on high accuracy.

# Iris data set

```{r}
data("iris")
str(iris)
```
# Scatter Plot & Correlations

```{r}
library(psych)
pairs.panels(iris[,-5],
             gap = 0,
             bg = c("red", "yellow", "blue")[iris$Species],
             pch=21)

```

# R package

```{r}
pc <- prcomp(iris[,-5],
             center = TRUE,
            scale. = TRUE)
attributes(pc)
```
```{r}
pc$center
```

```{r}
print(pc)
```
```{r}
summary(pc)
```

# Orthogonality of PCs

```{r}
pairs.panels(pc$x,
             gap=0,
             bg = c("red", "yellow", "blue")[iris$Species],
             pch=21)
```

```{r}
#calculate total variance explained by each principal component
var_explained = pc$sdev^2 / sum(pc$sdev^2)

#create scree plot
library(ggplot2)

qplot(c(1:4), var_explained) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0, 1)
```
- PC1 explains about 73.7% and PC2 explained about 22.1% of variability.

- Another way interpreting the plot is PC1 is positively correlated with the variables Petal Length, Petal Width, and Sepal Length, and PC1 is negatively correlated with Sepal Width.

- PC2 is highly negatively correlated with Sepal Width.

```{r}
library(ggplot2)
library(usethis)
library(devtools)
install_github("vqv/ggbiplot")
library(ggbiplot)
g <- ggbiplot(pc,
              obs.scale = 1,
              var.scale = 1,
              groups = iris$Species,
              ellipse = TRUE,
              circle = TRUE,
              ellipse.prob = 0.68)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal',
               legend.position = 'top')
print(g)
```

Bi plot is an important tool in PCA to understand what is going on in the dataset.



